Staged Abstract Interpreters

### 2019-09-15 paper/10_intro.tex
polish

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura projections. 
The first Futamura projection states that specializing an interpreter to an
input program removes interpretative overhead and yields an equivalent
executable that runs faster. Moreover, a procedure that can specialize an
interpreter to any input program is equivalent to a compiler.
In recent years, the idea of Futamura projections has attracted more attention,
and has been successfully applied to many real-world scenarios, such as
building JIT compilers for dynamic languages
\cite{Bolz:2009:TMP:1565824.1565827, Marr:2015:TVP:2814270.2814275} and building query
compilers \cite{DBLP:conf/sigmod/TahboubER18}.  The question here is can we
apply Futamura projections to an abstract interpreter and obtain a compiler
for abstract interpretation?
%apply Futamura projections to a general abstract interpreter to make it run faster? 
The solution is not obvious, as we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-points iteratively.

@@

The first Futamura projection shows that specializing an interpreter w.r.t. an
input program yields an equivalent program. For instance, assume that 
@eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization we
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied
% with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The earlier approach to realizing Futamura projections is partial
evaluation (PE) \cite{DBLP:books/daglib/0072559}.
Offline PE relies on a separate binding-time analysis (\textit{static} or
\textit{dynamic}), and online PE identifies the binding-times during the
specialization.
The underlying representation of binding-times can be viewed as a
two-level semantics \cite{NIELSON1989117, NIELSON198859,
Nielson:1992:TFL:130665} that distinguishes between compile-time computation
and run-time computation.  However, it is hard to precisely analyze
binding-times given an arbitrary program. As an alternative approach,
multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97}
allows programmers to annotate the binding-times of variables and control which
part of the program should be specialized.
At the syntactic level, traditional MSP languages (e.g., MetaML/MetaOCaml) use
quasi-quotations and escapes as annotations. However, the syntactic
quasi-quotation is a weak two-level semantics, and does not always preserve the
evaluation order of staged expressions.
%The MSP system will then check whether these
%annotations are consistent, and specialize the program using that information.
Besides the syntactic approach, staging annotations can also be purely
type-based.  One of the examples is the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala, which provides a stronger guarantee of
evaluation order by generating code in administrative normal form.

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract interpreter
as the unstaged baseline.  Similar to two-level semantics \cite{NIELSON1989117},
monads provide a good abstraction to hide the detail of abstract interpretation
semantics. However, monads do not have a clear stage distinction. We further
introduce binding-times and make the monads binding-time polymorphic. Our
proposed framework adopts type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projections of
an abstract interpreter for a stateful higher-order language. By deriving staged
monads that can be used to perform code generation, we obtain a staged abstract
interpreter that can be used for specialization. Through staging, the generated
code is specialized to the input program, and all the monadic operations are
inlined and compiled down to low-level Scala code.

>>>>

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura projections. 
The first Futamura projection states that specializing an interpreter to an
input program removes interpretative overhead and yields an equivalent
executable that runs faster. Moreover, a procedure that can specialize an
interpreter to any input program is equivalent to a compiler.
In recent years, the idea of Futamura projections has attracted more attention,
and has been successfully applied to many real-world scenarios, such as
building JIT compilers for dynamic languages
\cite{Bolz:2009:TMP:1565824.1565827, Marr:2015:TVP:2814270.2814275} and building query
compilers for databases \cite{DBLP:conf/sigmod/TahboubER18, DBLP:conf/osdi/EssertelTDBOR18}.
The question here is can we apply Futamura projections to an abstract
interpreter and obtain a compiler for abstract interpretation?
%apply Futamura projections to a general abstract interpreter to make it run faster? 
The solution is not obvious, as we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-points iteratively.

@@

The first Futamura projection shows that specializing an interpreter w.r.t. an
input program yields an equivalent program. For instance, assume that 
@eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e$@ : Expr@, by applying the specialization we
obtain a specialized program $\texttt{eval}_{\texttt{e}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied
% with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e ]\!] arg $.
The earlier approach to realizing Futamura projections is partial
evaluation (PE) \cite{DBLP:books/daglib/0072559}.
Offline PE relies on a separate binding-time analysis (\textit{static} or
\textit{dynamic}), and online PE identifies the binding-times during the
specialization.
The underlying representation of binding-times can be viewed as a
two-level semantics \cite{NIELSON1989117, NIELSON198859,
Nielson:1992:TFL:130665} that distinguishes between compile-time computation
and run-time computation.  However, it is hard to precisely analyze
binding-times given an arbitrary program. As an alternative approach,
multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97}
allows programmers to annotate the binding-times of variables and control which
part of the program should be specialized.
At the syntactic level, traditional MSP languages (e.g., MetaML/MetaOCaml) use
quasi-quotations and escapes as annotations. However, the syntactic
quasi-quotation is a weak two-level semantics, and does not always preserve the
evaluation order of staged expressions.
%The MSP system will then check whether these
%annotations are consistent, and specialize the program using that information.
Besides the syntactic approach, staging annotations can also be purely
type-based.  One of the examples is the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala, which as well provides a stronger
guarantee of preserving the evaluation order by generating code in
administrative normal form.

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract interpreter
as the unstaged baseline.  Similar to two-level semantics \cite{NIELSON1989117},
monads provide a good abstraction to hide the detail of abstract interpretation
semantics. However, monads do not have a clear stage distinction. We further
introduce binding-times and make the monads binding-time polymorphic. Our
proposed framework adopts type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the first Futamura projection of
an abstract interpreter for a stateful higher-order language. By deriving staged
monads that can be used to perform code generation, we obtain a staged abstract
interpreter that can be used for specialization. Through staging, the generated
code is specialized to the input program, and all the monadic operations are
inlined and compiled down to low-level Scala code.

### 2019-09-15 paper/20_prelim.tex
polish

Similar to Haskell's @do@-notation, Scala provides special syntactic support
for monadic operations through @for@-comprehension.  For example, an object of
@List[A]@ is an instance of the @List@ monad, with element type @A@.
To compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax:
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(3, 4)
  for { x <- xs; y <- ys } yield (x, y)   // List((1,3), (1,4), (2,3), (2,4))
\end{lstlisting}

The Scala compiler translates the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@. The last binding
in @for@-comprehensions is translated into a @map@ operation, where the argument of
@yield@ becomes the body expression inside that @map@ operation. The
bindings before the last one are all translated into calls of @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore, it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified versions of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad type as argument and produce another monad type. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the ones from Haskell.

@@

\paragraph{Basic Types} We start with some basic type definitions that are used
in the interpreter. The identifiers in the program are represented by strings.
To represent states, two required components for the interpreter are
environments, represented by type @Env@, and stores, represented by type
@Store@. An @Env@ maps identifiers to addresses, and a @Store@ maps addresses to
values. An environment captures bound variables in the scope of current control, and
a store models a persistent heap through the program run-time.  Currently,
the domains of addresses and values are still abstract; hence, they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

\paragraph{Binding-time Abstraction}
The binding-time type is declared as a higher-kinded type @R[_]@ \cite{Ofenbeck:2017:SGP:3136040.3136060}.
Now we can use @R@ to annotate other data-types in the interpreter;
it is also injected into the monadic type @MonadOps@.  If we instantiate
@R@ as the identity type (i.e., @type R[T] = T@), then the generic interpreter is a
standard definitional interpreter that will execute the program.  In Section
\ref{stagedinterp}, we instantiate @R@ using LMS's built-in next-stage
type annotation @Rep@, which makes the interpreter act as a compiler.

@@

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
types @Env@ and @Store@ are derived automatically. To ensure the freshness of
address allocations, we use type @Int@ and always return a number that is greater
than the size of the current store.
A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a $\lambda$-term and an environment. The final
result of the interpreter is a pair of values and stores.
To distinguish from the monadic values of type @Ans@ produced by the
interpreter, later we use term \textit{grounded values} to denote such final values.
The two elements in the type @Result@ are annotated by the binding-time @R@,
because they can be next-stage objects.  We also define a standard fixed-point
combinator to close the open-recursive function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int;  sealed trait Value
    case class IntV(i: Int) extends Value;  case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

>>>>

Similar to Haskell's @do@-notation, Scala provides special syntactic support
for monadic operations through @for@-comprehension.  For example, an object of
type @List[A]@ is an instance of the list monad, with element type @A@.
To compute the Cartesian product of two lists of numbers, we can use Scala's
@for@-comprehension syntax:
\begin{lstlisting}
  val xs = List(1, 2); val ys = List(3, 4); 
  for { x <- xs; y <- ys } yield (x, y)   // List((1,3), (1,4), (2,3), (2,4))
\end{lstlisting}

The Scala compiler translates the above @for@-comprehension expression into
an equivalent one using @flatMap@ and @map@. The last binding
of @for@-comprehensions is translated into a @map@ operation, where the argument of
@yield@ becomes the body expression inside that @map@ operation. The
bindings before the last one are all translated into @flatMap@:
\begin{lstlisting}
  xs.flatMap { case x => ys.map { case y => (x, y) } }
\end{lstlisting}

Note that here the monadic object of type @List[_]@ encapsulates the data internally.
Therefore, it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified versions of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad type as argument and produce another monad type. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the ones from Haskell.

@@

\paragraph{Basic Types} We start with some basic type definitions that are used
in the interpreter. The identifiers in the program are represented by strings.
To represent states, two required components for the interpreter are
environments, represented by type @Env@, and stores, represented by type
@Store@. An @Env@ maps identifiers to addresses, and a @Store@ maps addresses to
values. An environment captures bound variables in the scope of current control, and
a store models a persistent heap through the program run-time.  Currently,
the domains of addresses and values are still abstract; hence, they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value                 
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value] 
    type R[_] // Binding-time as a higher-kinded type
    ...  }
\end{lstlisting}

\paragraph{Binding-time Abstraction}
The binding-time type is declared as a higher-kinded type @R[_]@ \cite{Ofenbeck:2017:SGP:3136040.3136060}.
Now we can use @R@ to annotate other data types in the interpreter.
Later, the binding-time information is also injected into the monadic type
@MonadOps@.  If we instantiate @R@ as the identity type (i.e., @type R[T] = T@), 
then the generic interpreter is a standard definitional interpreter that
will execute the program.  In Section \ref{stagedinterp}, we instantiate @R@
using LMS's built-in next-stage type annotation @Rep@, which makes the
interpreter act as a compiler.

@@

\paragraph{Concrete Components}
The two types we need to concretize are addresses @Addr@ and values @Value@. The
types @Env@ and @Store@ are derived automatically. To ensure the freshness of
address allocations, we use type @Int@ and always return a number that is greater
than the size of the current store.
A value can be either a tagged number \texttt{IntV}, or a closure
\texttt{CloV} that contains a $\lambda$-term and an environment. The final
result of the interpreter is a @Value@ paired with a @Store@.
To distinguish from the monadic values of type @Ans@ produced by the
interpreter, later we use term \textit{grounded values} to denote such final values,
which can be obtained by running the monads.
The two elements in the type @Result@ are annotated by the binding-time @R@,
because they can be next-stage objects.  We also define a standard fixed-point
combinator to close the open-recursive function @ev@.
\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int;  sealed trait Value
    case class IntV(i: Int) extends Value;  case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}

### 2019-09-15 paper/30_stageconc.tex
polish

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} which
uses syntactic quotations, LMS distinguishes binding-times based on types. 
To achieve this, LMS provides a type constructor @Rep[T]@, where @T@ can be an
arbitrary type, indicating that a value of type @T@ will be known at a future
stage.  Operations acting on @Rep[T]@ expressions will be residualized as
generated code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

@@

In @close@, we first create a function @f@, which input and output are both
next-stage values and stores.  The function @f@ is a current-stage and we then
use @emit_complied_clo@ to denote it to a next-stage Scala function.
The method @emit_compiled_clo@ produces current-stage representations of
next-stage function values, i.e., of type @Rep[Value]@.
Inside function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also with the prepared environment and
store.  By doing so, we have \textit{collapsed} the monadic values of @AnsM@ to
grounded values within staging; therefore in the future stage, there will be no
monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

>>>>

Lightweight Modular Staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library. LMS enables
runtime code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14,
DBLP:conf/gpce/CalcagnoTHL03} which uses syntactic quotations, LMS
distinguishes binding-times based on types and preserves the evaluation order
via an internal ANF transformation of the staged program.
To achieve typed-based MSP, LMS provides a type constructor @Rep[T]@, where @T@
can be an arbitrary type, indicating that a value of type @T@ will be known at
a future stage.  Operations acting on @Rep[T]@ expressions will be residualized
as generated code.
A classic example for introducing multi-stage programming is the power function
that computes $b^x$. Usually, it is implemented as a recursive function:
\begin{lstlisting}
  def power(b: Int, x: Int): Int = if (x == 0) 1 else b * power(b, x - 1)
\end{lstlisting}

@@

In @close@, we first create a current-stage function @f@, whose input and
output are both next-stage values and stores.  We then use @emit_complied_clo@
to delay @f@ to a next-stage Scala function.
The method @emit_compiled_clo@ produces current-stage representations of
next-stage function values, i.e., of type @Rep[Value]@.
Inside function @f@, we can access the evaluator via @ev@.
However, @ev@ produces a monadic value of type @AnsM@, which can only exist at
the current stage. To connect current-stage monadic values and future-stage
grounded values, the evaluator @ev@ is supplied with not only the body
expression @e@ to be specialized, but also with the prepared environment and
store.  By doing this, we have \textit{collapsed} the monadic values of type
@AnsM@ to grounded values (of type @Rep[ValSt]@) within staging; therefore in
the future stage, there will be no monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

### 2019-09-15 paper/40_absinterp.tex
space
polish

Many abstract domains in abstract interpreters can be formulated by complete
lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage-polymorphic lattices. Similar to the @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to the
trait for annotating the binding-times, where @R[_]@ wraps the data type
@S@ and @Boolean@ in these operations.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
  \begin{lstlisting}[style=small]
  trait SPLattice[S, R[_]] {
    val ⊤: R[S];  val ⊥: R[S]
    def ⊑(l1: R[S], l2: R[S]): R[Boolean]
    def ⊔(l1: R[S], l2: R[S]): R[S]
    def ⊓(l1: R[S], l2: R[S]): R[S]
  }
  trait Lattice[S] extends SPLattice[S, NoRep]
  \end{lstlisting}
  \caption{trait \texttt{SPLattice} and \texttt{Lattice}} \label{fig:splattice}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.6\textwidth}
\begin{lstlisting}[style=small]
  def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
    lazy val ⊤: Set[T] = error("No representation for ⊤")
    lazy val ⊥: Set[T] = Set[T]()
    def ⊑(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf  l2
    def ⊔(l1: Set[T], l2: Set[T]): Set[T]  = l1 union     l2
    def ⊓(l1: Set[T], l2: Set[T]): Set[T]  = l1 intersect l2
  }
\end{lstlisting}
  \caption{The power set instance for lattice} \label{fig:powerset}
\end{subfigure}
\end{figure}

@@

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore both of the branches.
Also, at some later point, we need to compute the join of two paths.  Another
source of nondeterminism comes from function applications, for instance,
consider an application @f(a)@, the possible $\lambda$-terms of @f@ are
computed approximately.  As the result, we need to store all the possible
callees, as well as we will retrieve a set of closures when dereferencing an
address.

@@

This monad stack scheme (@AnsM[T]@) leads to a different type of grounded
values. Recall that in the concrete setting, we only have a reader monad (for
environments) and a state monad (for stores). The reader effect is not
persistent to the final result. Hence, together with the value produced by the
interpreter, the final result type is just a pair of @Value@ and @Store@.
However, under the new monad stack for abstract interpretation, we have a
@SetT@ monad transformer inside the environment monad and store monad. Therefore, the type
@Set@ becomes the container type of the pairs of values and stores, i.e.,
@Set[(Value, Store)]@.  Also, note that the reader and state monad for the
caches both inhibit internally of the nondeterminism monad; as the result, the
final result type pairs the set of value-stores with the cache type from
that state monad, as shown below.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed according
to the new monad stack scheme and value domains. One of the notable changes is
that the store update operator merges the new values with existing values.  As
we mentioned before, sometimes we have to explore both of the branches for
conditionals: in method @br0@, we combine the results using the @mplus@
operation from @MonadPlus@, which requires the value domains are
join-semilattices (in our case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument of @ap_clo@ is a set of closures
@funs@, so we cannot directly destruct it with a syntactic @Lam@ by pattern
matching. Instead, we use an auxiliary function @lift_nd@ that takes a set and
lifts the elements in the set into the monad stack. Then we can
straightforwardly implement the method @ap_clo@ still in monadic style, where the
closures come from the monads and thus the nondeterminism can be naturally
handled. The light gray line shows what is added from its concrete counterpart.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

\paragraph{Caching and Fixpoint Iteration}
As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called a \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use an @in@ cache and an @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and the @out@ cache is constantly updated during the current iteration. In
the next iteration, the last @out@ cache is served as the @in@ cache, and
an empty cache is plugged-in to the @out@ slot. Once the @out@ cache does not
contain any new information compared with the @in@ cache, the fixed-point is
reached.
In our monad stack, the @in@ and @out@ caches are modeled by the reader monad and
state monad, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

@@

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact has been discovered.
Therefore, the iteration should end, and we have reached the least fixed-point.
\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ0)(σ0)(cache0)(cache0).run
\end{lstlisting}

>>>>

Many abstract domains in abstract interpreters can be formulated as complete
lattices.
A complete lattice over set $S$ is a tuple $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage-polymorphic lattices. Similar to the @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to the
trait for annotating the binding-times, where @R[_]@ wraps the data type
@S@ and @Boolean@ in these operations.

\begin{figure}[h!]
  \centering
  \vspace{-1em}
  \begin{subfigure}[b]{0.45\textwidth}
  \begin{lstlisting}[style=small]
  trait SPLattice[S, R[_]] {
    val ⊤: R[S];  val ⊥: R[S]
    def ⊑(l1: R[S], l2: R[S]): R[Boolean]
    def ⊔(l1: R[S], l2: R[S]): R[S]
    def ⊓(l1: R[S], l2: R[S]): R[S]
  }
  trait Lattice[S] extends SPLattice[S, NoRep]
  \end{lstlisting}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.6\textwidth}
  \begin{lstlisting}[style=small]
    def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
      lazy val ⊤: Set[T] = error("No representation for ⊤")
      lazy val ⊥: Set[T] = Set[T]()
      def ⊑(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf  l2
      def ⊔(l1: Set[T], l2: Set[T]): Set[T]  = l1 union     l2
      def ⊓(l1: Set[T], l2: Set[T]): Set[T]  = l1 intersect l2
    }
  \end{lstlisting}
  \end{subfigure} 
  \vspace{-1.5em}
  \caption{trait \texttt{SPLattice} and \texttt{Lattice} (left),
  and the power set instance (right) } \label{fig:splattice}
  \label{fig:powerset}
\end{figure}
\vspace{-1em}

@@

It is important that the abstract stores now map addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
runtime values.  This can be justified by the approximation nature and
nondeterminism during the analysis.  For example, when analyzing a
conditional expression, we may not have enough information to decide which
branch will be taken, thus a sound treatment is to explore both of the branches.
Also, at some later point, we need to compute the join of two paths.  Another
source of nondeterminism comes from the abstract semantics of function
application. For instance, consider an expression @f(a)@, if there are multiple
possible target closures of @f@, then we have to apply all of them
nondeterminitically.
%As the result, we need to store all the possible callees, as well as we will
%retrieve a set of closures when dereferencing an address.

@@

This monad stack scheme (@AnsM[T]@) leads to a different type of grounded
values. Recall that in the concrete setting, we only have a reader monad (for
environments) and a state monad (for stores). 
Different from state effects, reader effects are not visible at the final
result. Hence, together with the value produced by the interpreter, the final
result type is just a pair of @Value@ and @Store@.  However, under the new
monad stack for abstract interpretation, we have a @SetT@ monad transformer
inside the environment monad and store monad. Therefore, the type @Set@ becomes
the container type of the pairs of values and stores, i.e., @Set[(Value, Store)]@.
Also, note that the reader and state monad for the caches both
inhabit the nondeterminism monad; as the result, the final result
type pairs the set of value-stores with the cache type from that state monad,
as shown below.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed according
to the new monad stack scheme and value domains. One of the notable changes is
that the store update operator merges the new values with existing values.  As
we mentioned before, sometimes we have to explore both of the branches for
conditionals: in method @br0@, we combine the results using the @mplus@
operation from @MonadPlus@, which requires the value domains to be
join-semilattices (in our case, @Set@ and @Map@).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is that now the first argument of @ap_clo@ is a set of closures
@funs@, so we cannot directly match it against with a syntactic @Lam@ by pattern
matching. Instead, we use an auxiliary function @lift_nd@ that takes a set and
lifts the elements in the set into the monad stack. Then we can
straightforwardly implement the function @ap_clo@ still in monadic style, where the
closures come from the monads and thus the nondeterminism can be naturally
handled. The light gray line shows what is added from its concrete counterpart.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(funs: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](funs)}!
    α <- alloc(x)
    _ <- set_store(α → funs)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

\paragraph{Caching and Fixpoint Iteration}
As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination.
The caching algorithm is also called a \textit{co-inductive} caching or
\textit{truncated depth-first evaluation} \cite{Rosendahl:AbsIntPL}. It has
been used in other abstract interpreters or fixed-point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use an @in@ cache and an @out@ cache during
the depth-first evaluation. The @in@ cache stores the result from the last
iteration, and the @out@ cache is constantly updated during the current iteration. In
the next iteration, the last @out@ cache is used as the @in@ cache, and
an empty cache is plugged into the @out@ slot. Once the @out@ cache does not
contain any new information compared with the @in@ cache, the fixed-point is
reached.
In our monad stack, the @in@ and @out@ caches are modeled by the reader monad and
state monad, respectively. We first define several methods to manipulate
the two caches through the monad stack (the implementations are elided).
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

@@

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact have been discovered.
Therefore, the iteration should end, and we have reached the least fixed-point.
\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ0)(σ0)(cache0)(cache0).run
\end{lstlisting}
